---
title: "final_R_reference"
author: "kato"
date: "2022-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# R Reference Guide {.tabset .tabset-pills}


## Loading Data and Packages
When attempting to use add-on packages  that are not a part of the base R package, you must notify R that you want to use them using require() and library()

```{r load-in}
library("palmerpenguins")
require(here)
```
Now the "palmerpenguins" data set and the here() command are ready to use!

If you would like to load in a csv file that is on your computer you can use read.csv() to choose the file and here() to bring it into your R workspace
```{r outside-files}
ginkgo = read.csv(here("data", "ginkgo_data_2022.csv"))
```
This creates a data frame with the data you loaded in.]

## Data Structures
The function c() combines or concatenates its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements). All of the elements must be of the same type. I can’t combine character and numeric types in the same call to c()

```{r c}
## Create a vector of numbers:
num_vec  = c(1, 5, 8, 9, 13, 57)

## Create a vector of characters:
char_vec = c("a", "spiderz", "data is le epic swag")
```

The length() function is used to tell you the length/sum of items within a vector. 
```{r length}
## If I want to know how many items are in my vector of numbers I could simply:
 length(num_vec)

## I can do the same with my character vector:
length (char_vec)
```

matrix() creates a matrix from a given set of values. 
* data = the data used to fill in the matric
* nrow = # of rows
* ncol = # of columns
* byrow (TRUE/FALSE) = whether or not you want the matric to be sorted by row (true) or column(false)
```{r matrix}
num_col = matrix(data = num_vec, nrow = 2, ncol = 3, byrow = FALSE)
print(num_col)

num_row = matrix(data = num_vec, nrow = 2, ncol = 3, byrow = TRUE)
print(num_row)
```
Notice the difference of the organization?


data.frame() creates a table that display many different groups of data. Different columns can house different types of data but very column must have the same type of data. 
For example, lets say we work at a veterinarians office and we need to make a table for the patients we saw that day: 
Column 1 could be all characters: like the species of dog, Column 2 could be logical: like whether or not the dog had fleas, and column 3 for be numbers: like the dogs weight in pounds.

The structure is as follows:
```{r data.frame}
dog_data <- data.frame(
  dog_species = c("German Shephard","Boxer","Rottweiler","Yorkie","Jack Russel"),
  fleas_present = c(TRUE,FALSE, FALSE, TRUE, FALSE),
  dog_weight_lbs = c(80.4, 75.6, 100.5, 15.3, 20.6 )
)

print(dog_data)
```

Now that you have created a data frame, it helps to be able to understand the dimensions of it. You can do that with the following functions: 
```{r nrow}
## to find the number of rows in a data set you can:
nrow(ginkgo)
## to find the number of columns in a data set you can:
ncol(ginkgo)
## and if you want to know both the row and column dimensions (IN THAT ORDER!) you can:
dim(ginkgo)
```
## Subsetting 
To retrieve a specific column from your data set you can call the name of the data frame and use the "$" symbol to specify the column you wish to view.This will print the data from that column as such:
```{r ginkgo site_id}
ginkgo$site_id
```

If you want to select a specific row/column based on their order you can use []. The format is as such: The first number represents the row number, and the second number represents the column number. If you wanted data from the first row and second column you would write [1,2]. If you only want data from a row or column, you can leave the other piece blank. Here is how you apply it:
```{r rows and columns}
## we want ONLY row 1
ginkgo[1,]

## we want row 2, column 3
ginkgo[2,3]

## we want ONLY column 3 
ginkgo[,3]
```

To create a new data frame that applies to a certain criteria, you can use subset(). In this case, if I only wanted to look at data from Adelie penguins, I could drop any data that was associated with the other species.
```{r subset()}
dat_adelie = droplevels(subset(penguins, species == "Adelie"))
```
## Numerical Data Exploration
If we want to create a summary of a data set that includes information such as: the min. and max., the quartiles, the mean, and the median we can use: summary(). Here is a summary of the ginkgo data.
```{r ginkgo summary}
## this creates a summary of the entire data frame
summary(ginkgo)
```


If you want to find the mean of a group of data you can use: mean()
```{r mean}
## The mean maximum width of the ginkgo leaves is:
mean(ginkgo$max_width)
```

We can execute something similar to find the standard deviation of a group of data with: sd()
```{r sd}
## The standard deviation of maximum width of the ginkgo leaves is:
sd(ginkgo$max_width)
```

## Graphical Data Exploration 
In order to visualize data, youll need to graph it. We can start out with a basic scatter plot using plot(). Lets plot max leaf depth (x) and max leaf width (y).
```{r ginkgo}
## x is max leaf depth, y is max leaf width
plot(x = ginkgo$max_depth, y = ginkgo$max_width)
```


Now this is a cool graph, but it is missing a lot of crucial details and personalizations like a title, axis labels, etc. Let's fix that!
```{r new-ginkgo}
## x is max leaf depth, y is max leaf width
plot(x = ginkgo$max_depth, y = ginkgo$max_width,
     ## the color of the data points. Enter your chosen color in parentheses or the hex code. 
      col = "orange",
     ## the shape of the data points. The default is unshaded circles. I want mine to be triangles, so I will find the corresponding symbol and include it
      pch = 17,
     ## refers to the relative size of the text. The default is '1', and whatever you enter will be a permutation of that. I want the text to be 80% of the original size so I will do the following:
      cex = .8,
     ## Dictates the title of the graph. Always put it in quotation marks and use \n to create a new line
      main = "Maximim leaf depth and width of \n Ginkgo leaves found on UMass Campus", 
     ## dictates the x axis label. our x axis is max depth
      xlab = "Maximum Leaf Depth",
     ## dictates the y axis label. our y axis is max width
      ylab = "Maximum Leaf Width",
     ## Dictates the size of the X-axis. If I want to extend the x-axiz past the default I can do so with this notation: range(min value, max value)
      xlim = range(0,140),
     ## Same as xlim, but on the y-axis
      ylim = range(0,160)
       )
```


If we wanted to make a histogram we could do this:
```{r hist-penguin}
hist(penguins$flipper_length_mm)
```

Again, this is a bit lackluster. Lets personalize this
```{r new-hist}
hist(penguins$flipper_length_mm,
  ## we know these from the last one:
  col = "purple",
  cex = .8,
  main = "penguin flipper length",
  xlab = "Flipper Length",
  ## the "breaks" arguments dictates the bin organization of the histogram. I am going to make the bins far too big to illustrate the usage!
  breaks =c(160,190, 220, 240))
```

Now if we wanted to make a boxplot, we could do this: 
```{r bonx-plot}
## blox plot of petiole length
boxplot(ginkgo$petiole_length, main = "Total Petiole Length", ylab = "Petiole Length")
## boxplot of petiole length conditioned on whether or not seeds were present on the tree
boxplot(petiole_length ~ seeds_present, dat = ginkgo, main ="Petiole Length by seed presence", cex = .6, ylab = "Petiole Length", xlab = "Seeds Present?")

```

If we wanted to create a panel of multiple graphs, we can using this method:
```{r 4panels}
## to put the graphs on a 4 graph panel in a 2/2 grid, we can use par(mfrow = c(# of rows, # of columns)). In this case I want 2 rows and 2 columns
par(mfrow = c(2,2))
hist(penguins$bill_length_mm, main = "Penguin Bill Length", xlab = "Bill Legnth (mm)")
hist(penguins$flipper_length_mm, main = "Penguin Flipper Length",  xlab = "Flipper Legnth (mm)")
hist(penguins$bill_depth_mm, main = "Penguin Bill Depth",  xlab = "Bill Depth (mm)")
hist(penguins$body_mass_g, main = "Penguin Body Mass",  xlab = "Body Mass (g)")

```


## Distribution Functions


The function dnorm() gives you the value of the **probability density function** (pdf) for a normal distribution with certain parameters (Either a value or a vector of values) 
* x = your value 
* mean = population mean
* sd = the population standard deviation.

Lets start with a standard normal distribution:

```{r dnorm}
dnorm(x = 0, mean = 0, sd = 1)
```

pnorm() gives you the value of the **cumulative density function** (cdf) of the normal distribution given certain parameters (Either a value or a vector of values) 
* q = your value
* mean = population mean
* sd = the population standard deviation
* lowertail = TRUE/FALSE depending on if you want the calculation to estimate greater than the quartile (false) or less than the quartile (true). It is FALSE by default. 

pnorm() gives you the area to the **left** of a given value x in the normal distribution, unless told otherwise!

Lets try again with a standard normal distribution:

```{r pnorm }
pnorm(q = .5, mean = 0, sd = 1, lower.tail = FALSE)
```

This function gives you the **inverse cumulative density function** (cdf) of the normal distribution given a certain parameters
* p = your value
* mean = the population mean
* sd = standard deviation 
* lowertail = TRUE/FALSE depending on if you want the calculation to estimate greater than the quartile (false) or less than the quartile (true). It is FALSE by default. 

Lets try again with a standard normal distribution:
```{r qnorm}
qnorm(q = .5, mean = 0, sd = 1, lower.tail = FALSE)
```

dbinom() gives you the value of the **probability density function** (pdf) of the binomial distribution given certain paramters: 
* x = your number of successes
* size = # of trials
* prob = probability of success

whats the probability of getting 5 heads when I flip a coin 8 times?
```{r dbinom}
## 5 heads = 5 successes -> 8 coin flips = 8 trials -> flipping coin = prob. of .5
dbinom(x = 5, size = 8, prob = .5)
```

pbinom() gives you the value of the **cumulative density function** (cdf) of the binomial distribution given certain parameters:
* q = your value
* size = number of trials 
* prob = probability of success 
* lowertail = TRUE/FALSE depending on if you want the calculation to estimate greater than the quartile (false) or less than the quartile (true). It is FALSE by default. 


pbinom gives you the area to the **left** of your q value in the binomial distribution, unless otherwise told to!


For example: find the probability of **more than** 3 heads during 5 coin flips 
```{r pbinom}

pbinom(q = 3, size=5, prob=.5, lower.tail=FALSE)
```

qbinom() gives you the value of the **inverse cumulative density function** (cdf) of the binomial distribution given certain parameters:
* q = your value
* size = number of trials 
* prob = probability of success 
* lowertail = TRUE/FALSE depending on if you want the calculation to estimate greater than the quartile (false) or less than the quartile (true). It is FALSE by default. 

qbinom gives you the area to the **left** of your q value in the binomial distribution, unless otherwise told to!

find the 15th quantile out of 100 trials with a probability of 0.5 of success for each trial
```{r qbinom}

qbinom(.15, size=100, prob=.5)
```


